---
title: "Exercise B"
output: html_notebook
---
# Lecture 2, Exercise B
## No work from book

First, load libraries.
```{r}
library(nycflights13)
library(tidyverse)
```

Look at data.
```{r}
flights
count(flights)
```

Filter the data. Here, all flights in month 1. Or combine conditions
```{r}
filter(flights, month == 1)
filter(flights, month == 1, day == 1)
```

## Identifying Missing Data

You can also simply remove the rows with missing values with the function “filter()”. Just negate the “is,na()” test, i.e. select all data lines which DON’T contain a missing value.

```{r}
filter(flights, ! is.na(dep_time))
count(filter(flights, ! is.na(dep_time)))
```

## Save cleaned data

Make a variable to save the data.
```{r}
my_flights <- filter(flights, ! is.na(dep_time))
```

## Replace missing data

Here, use the scheduled departure time as the departure time, when there is no depature time.

```{r}
flights_with_replaced_dep_time <- flights
flights_with_replaced_dep_time$dep_time <- ifelse(is.na(flights$dep_time), flights$sched_dep_time, flights$dep_time)
```

Check the work with an example

```{r}
filter(flights, tailnum == "N18120")
filter(flights_with_replaced_dep_time, tailnum =="N18120")
```

You can also replace NA with other values. Here, 12:00 (just because)

```{r}
replacement <- 1200

flights_with_replaced_dep_time <-
  ifelse(is.na(flights$dep_time),
         replacement,
         flights$dep_time)
```

Or replace it with the mean.
```{r}
replacement <- as.integer(mean(flights$dep_time, na.rm = TRUE))

replacement

flights_with_replaced_dep_time <-
  ifelse(is.na(flights$dep_time),
         replacement,
         flights$dep_time)
```
# Identifying and Eliminating Outliers

Analyse departure delay
```{r}
ggplot(flights) +
  geom_point(mapping = aes(x = flight, y = dep_delay))
```

There are some flights that depart earlier than scheduled. 
```{r}
arrange(flights, dep_delay)
```
We catch them.
```{r}
minus_delay <- filter(flights, dep_delay <= 0)
```

Analyze the distribution of the negative departure delay:
```{r}
boxplot(minus_delay$dep_delay)
```

Some values are out of range. We remove all lines with a lower negative departure delay than 29.
```{r}
my_flights <- filter(flights, dep_delay > -29)
```

