---
title: "Exercise D"
output: html_notebook
---
```{r}
library(fastDummies)
```

# Qualitative Data

```{r}
titanic <- data.frame(Titanic)
titanic
my_titanic <- titanic
```

# Categorical Variables

## Tranforming categorical variable Survived

First, turn "survived" into a factor. It contains "No" and "Yes" and internally is already an integer.
```{r}
f <- factor(titanic$Survived)
levels(f)
typeof(f)
```

Now, we need to transform f to an integer and write it to the column/variable "Survived"
```{r}
my_titanic$Survived <- as.integer(f)
```

Or everything on one line:
```{r}
my_titanic$Survived <- as.integer(factor(titanic$Survived))
my_titanic
```
Note: "Survived" == 1 if the person did not survive. Which seems weird! Survived == 2 if the person did survive.

## Tranforming the categorical variable Sex

This variable is NOT ordered. So it cannot be transformed into a factor. Instead, we make a boolean for each value.

"dummy_cols" selects the column "Sex", removes it, and adds for each value a new (0/1) column. So two columns here: Sex_Male and Sex_Female.
```{r}
my_titanic <- dummy_cols(my_titanic, select_columns = "Sex", remove_selected_columns = TRUE)
my_titanic
```

# Ordinal Variable

## Transforming the ordinal variable age

Age in this database is not a number, it is qualitative ("Child" or "Adult").
We just want to divide into children and adults, with Child = 1 and Adult = 2

```{r}
ordered(my_titanic$Age, levels = c("Child", "Adult"))
```

Convert it to integers
```{r}
as.integer(ordered(my_titanic$Age, levels = c("Child", "Adult")))
```
Now, we can replace the values in column Age
```{r}
my_titanic$Age <- as.integer(ordered(my_titanic$Age, levels = c("Child", "Adult")))
my_titanic
```
## Tranforming the partly ordinal variable Class

Class is qualitative but partially ordered (1st, 2nd, 3rd Class, BUT Crew). Therefore, we create a Boolean for Crew but ordered column for the other values.
```{r}
ifelse(my_titanic$Class == "Class",1,0)

my_titanic$Class_Crew <- ifelse(my_titanic$Class == "Crew",1,0)
```

Now order Class:

```{r}
ordered(my_titanic$Class, levels = c("Crew", "3rd", "2nd", "1st"))
```

Converting it into integers:
```{r}
as.integer(ordered(my_titanic$Class, levels = c("Crew", "3rd", "2nd", "1st")))
```
Since this includes Crew still, we need to get the "right" Class numbers by subtracting 1
```{r}
as.integer(ordered(my_titanic$Class, levels = c("Crew", "3rd", "2nd", "1st")))-1
```

The resulting dataset:
```{r}
my_titanic
```

