---
title: "Exercise C"
output: html_notebook
---

```{r}
library(nycflights13)
library(tidyverse)
```

## Deleting columns

Subset - the part you want to leave out.
```{r}
flights
my_flights <- subset(flights,select=-dep_delay)
my_flights

myflights <- subset(flights,select=-c(dep_delay, flight))
myflights
```

## Transforming departure delay

First remove all rows with missing values and remove rows with extreme negative delay.

```{r}
my_flights <- filter(flights, ! is.na(dep_time))
my_flights <- filter(my_flights, dep_delay > -29)

hist(my_flights$dep_delay)
```

This we can transform into a logarithmic distribution, to be more uniform

```{r}
hist(log(my_flights$dep_delay))


ggplot(
  data = my_flights,
  mapping = aes(x = log(dep_delay))
) +
  geom_histogram()
```

We produce NA for negative delay. To remove negative values without deleting them, we shift the delays by the most negative value, so all values are positive.

```{r}
minimum <- min(my_flights$dep_delay, na.rm = TRUE)
hist(log(my_flights$dep_delay - minimum))
```

## Normalizing Departure Time

First, remove missing values (flights that didn't depart)
```{r}
my_flights <- filter(flights, ! is.na(dep_time))
```

Apply min-max normalization to "dep_time" (range of 0000-2359, as in a day)
New min is 0 and new max is 1.
Then [(v-0000)/(2359-0000)] * (1 - 0) + 0 = v / 2359

*NOTE: This is bad code as explained just below it!*

```{r}
my_flights$dep_time <- my_flights$dep_time / 2359
my_flights
```

However, the coding of time in integer is not continuous. That is, 1178 does not exist.
We need to write a conversion function to translate to continuous numbers.

```{r}
time_conversion <- function(x) {
  h <- trunc(x/100,0)
  m <- x-(h*100)
  r <- m+(h*60)
  return(r)
}

my_flights$dep_time <- time_conversion(my_flights$dep_time) / (24*60)
my_flights
```

