---
title: "Data Science Lecture 2, Exercise A"
output: html_notebook
---

First, load ggplot2 to make pretty graphs. Corrplot used later in exercise.

```{r}
library(ggplot2)
library(corrplot)
```

We will work with mtcars dataset, so what does it look like?

```{r}
head(mtcars)
tail(mtcars)
```

Let's get our first insights into the dataset

```{r}
str(mtcars)
```

Compute the mean, median, and mode of column wt

```{r}
mean(mtcars$wt)
```

```{r}
median(mtcars$wt)
```

For mode, there is no built-in mode function! Create a variable for a table, then see what the one used the most is.

```{r}
y <- table(mtcars$wt)
y
names(y)[which(y==max(y))]
```

summary() also works well to do all of this.

```{r}
summary(mtcars)
```

Plot the column wt

```{r}
ggplot(mtcars) +
  geom_point(mapping = aes(x = 1:length(wt), y = wt))
```

Plot wt as function of drat

```{r}
ggplot(mtcars, aes(x = drat, y = wt)) +
  geom_point()
```

Boxplot wt

```{r}
boxplot(mtcars$wt)
```

Histogram and read ?hist for help

```{r}
hist(mtcars$wt)
```

```{r}
?hist
```

Q-Q Plot of wt The Q-Q plot, or quantile-quantile plot, is a graphical tool to help us assess if a set of data plausibly came from some theoretical distribution such as a Normal or exponential. If both sets of quantiles came from the same distribution, we should see the points forming a line that's roughly straight.

In R, there are two functions to create Q-Q plots: qqnorm and qqplot. qqnorm creates a Normal Q-Q plot. You give it a vector of data and R plots the data in sorted order versus quantiles from a standard Normal distribution. The qqplot function allows you to create a Q-Q plot for any distribution. Unlike the qqnorm function, you have to provide two arguments: the first set of data and the second set of data.

```{r}
qqnorm(mtcars$wt)
```

We can draw a line on our Q-Q Plot to help see. First build the plot, then draw the line.

```{r}
qqnorm(mtcars$wt)
qqline(mtcars$wt)
```

What correlation do columns have? Rounded to two decimals

```{r}
round(cor(mtcars),2)
```

Use corrplot to check the correlation (must be installed first).

```{r}

corrplot(cor(mtcars), type = "upper", order = "hclust", tl.col = "black", tl.srt = 45)
```
