---
title: "Lecture 3 - Linear Regression Exercises"
output: html_notebook
---
First, load the file
```{r}
Advertising <- read.csv("./Advertising.csv")
```

Summarize it
```{r}
dim(Advertising)
summary(Advertising)
```

Is there a linear relationship?
```{r}
plot(Advertising$TV, Advertising$sales)
```

We can do a least squares fit pretty easily
```{r}
lm.fit <- lm(Advertising$sales~Advertising$TV)
```

By attaching the dataset, we can call the columns directly.
```{r}
attach(Advertising)
lm.fit <- lm(sales~TV)
summary(lm.fit)
```
We can use the function confint() to find the confidence interval for 95%
```{r}
confint(lm.fit)
```
## Hypothesis testing

p-value: probability of observing a value equal or larger than t

If p-value is small enough (alpha) is less than our cutoff (5% or 1%), we can reject the null hypothesis and accept that there is a relationship.

Need high t-value (absolute value) and low p-value

## R-squared and RSE

This checks how good your line fits the data.
Want RSE to be low, R-squared to be high.

RSE is a measure of units of Y, so not always clear what value it *should* be.

R-squared is deviation from the mean. A high value (closer to 1), means the x-variable explains that proportion of the change in Y. A zero value means the X does not explain any change in Y.

# Multiple Regression

Multiple variables

```{r}
attach(Advertising)
lm.fit <- lm(data = Advertising, formula = sales~TV+radio+newspaper) # 3 predictors
lm.fit_multi <- lm(sales~., data = Advertising) # short-hand to use all predictors
lm.fit_single <- lm(sales~newspaper)

summary(lm.fit_multi)
summary(lm.fit_single)
```

Building three models by adding variables
```{r}
model1 <- lm(formula = sales ~ TV, data = Advertising)
summary(model1)

model2 <- lm(formula = sales ~ TV + radio, data = Advertising)
summary(model2)

model3 <- lm(formula = sales ~ ., data = Advertising)
summary(model3)
```
# Extensions of the Linear Model

## Variables Interact

Maybe two variables interact to increase the effect (example: two medications taken together, TV and radio ads at the same time)

We can see that interaction

```{r}
lm.fit1 <- lm(formula = sales ~TV*radio, data = Advertising) # we use * between the variables so R considers them independently as well as together
summary(lm.fit1)
```

## Non-Linear Relationships

