---
title: "Lecture 5 Exercises"
output: html_notebook
---

Resampling Methods

# Exercise 1 - The Validation Set Approach

```{r}
library(ISLR)
library(boot)
```

The Auto data set is contained in the ISLR package
```{r}
str(Auto)
attach(Auto)
```
Define training and test data
```{r}
set.seed(1)
(train=sample(nrow(Auto)/2)) # generate indices of a training data set by random sampling half of the observation indices of Auto
AutoTrain <- Auto[train,] # training data
AutoTest<- Auto[-train,] # test data
```
Fit 3 different regression models to the training data: Which model gives a better fit?

```{r}
(lm.fit <- lm(mpg~horsepower, data = AutoTrain)) # linear regression
(lm.fit2 <- lm(mpg~poly(horsepower,2), data = AutoTrain)) # quadratic regression
(lm.fit3 <- lm(mpg~poly(horsepower,3), data = AutoTrain)) # cubic regression
```

Pot the fits
```{r}
plot(mpg~horsepower, data = AutoTrain)
abline(lm.fit$coefficients, col="blue4") # plot linear fit

x <- with(Auto, seq(min(horsepower), max(horsepower), length.out = 2000)) # define x values to plot polynomial fits

y2 <- predict(lm.fit2, newdata = data.frame(horsepower=x)) # corresponding predicted values
lines(x, y2, col = "red4")

y3 <- predict(lm.fit3, newdata = data.frame(horsepower = x)) # corresponding predicted values
lines(x, y3, col = "green4")
```

Looking at the plots, I think the quadratic (red line) will evaluate the best because it turns upwards less as horsepower increases.

## Calculate the test MSEs of the 3 models (mean of squared errors on test set)
```{r}
(MSE <- mean((mpg-predict(lm.fit,Auto))[-train]^2)) 
(MSE2 <- mean((mpg-predict(lm.fit2,Auto))[-train]^2)) 
(MSE3 <- mean((mpg-predict(lm.fit3,Auto))[-train]^2))
```
I was right that fit2 is the best.

## Resample with a different training set
To get a different training set, we change the seed.

```{r}
set.seed(2)
(train=sample(nrow(Auto),nrow(Auto)/2)) # generate indices of a training data set by random sampling half of the observation indices of Auto 
AutoTrain <- Auto[train,] # training data 
AutoTest <- Auto[-train,] # test data

(lm.fit = lm(mpg~horsepower,data=AutoTrain)) # linear regression
(lm.fit2 = lm(mpg~poly(horsepower,2),data=AutoTrain)) # quadratic regression
(lm.fit3=lm(mpg~poly(horsepower,3),data=Auto,subset=train)) # cubic regression

(MSE <- mean((mpg-predict(lm.fit,Auto))[-train]^2)) 
(MSE2 <- mean((mpg-predict(lm.fit2,Auto))[-train]^2)) 
(MSE3 <- mean((mpg-predict(lm.fit3,Auto))[-train]^2))

```

```{r}
set.seed(3)
(train=sample(nrow(Auto),nrow(Auto)/2)) # generate indices of a training data set by random sampling half of the observation indices of Auto 
AutoTrain <- Auto[train,] # training data 
AutoTest <- Auto[-train,] # test data

(lm.fit = lm(mpg~horsepower,data=AutoTrain)) # linear regression
(lm.fit2 = lm(mpg~poly(horsepower,2),data=AutoTrain)) # quadratic regression
(lm.fit3=lm(mpg~poly(horsepower,3),data=Auto,subset=train)) # cubic regression

(MSE <- mean((mpg-predict(lm.fit,Auto))[-train]^2)) 
(MSE2 <- mean((mpg-predict(lm.fit2,Auto))[-train]^2)) 
(MSE3 <- mean((mpg-predict(lm.fit3,Auto))[-train]^2))

```

```{r}
set.seed(4)
(train=sample(nrow(Auto),nrow(Auto)/2)) # generate indices of a training data set by random sampling half of the observation indices of Auto 
AutoTrain <- Auto[train,] # training data 
AutoTest <- Auto[-train,] # test data

(lm.fit = lm(mpg~horsepower,data=AutoTrain)) # linear regression
(lm.fit2 = lm(mpg~poly(horsepower,2),data=AutoTrain)) # quadratic regression
(lm.fit3=lm(mpg~poly(horsepower,3),data=Auto,subset=train)) # cubic regression

(MSE <- mean((mpg-predict(lm.fit,Auto))[-train]^2)) 
(MSE2 <- mean((mpg-predict(lm.fit2,Auto))[-train]^2)) 
(MSE3 <- mean((mpg-predict(lm.fit3,Auto))[-train]^2))

```

```{r}
set.seed(5)
(train=sample(nrow(Auto),nrow(Auto)/2)) # generate indices of a training data set by random sampling half of the observation indices of Auto 
AutoTrain <- Auto[train,] # training data 
AutoTest <- Auto[-train,] # test data

(lm.fit = lm(mpg~horsepower,data=AutoTrain)) # linear regression
(lm.fit2 = lm(mpg~poly(horsepower,2),data=AutoTrain)) # quadratic regression
(lm.fit3=lm(mpg~poly(horsepower,3),data=Auto,subset=train)) # cubic regression

(MSE <- mean((mpg-predict(lm.fit,Auto))[-train]^2)) 
(MSE2 <- mean((mpg-predict(lm.fit2,Auto))[-train]^2)) 
(MSE3 <- mean((mpg-predict(lm.fit3,Auto))[-train]^2))

```

fit2 and 3 were always close, and it wasn't the case that fit2 always won.

# Exercise 1.1

Now lets draw a nice graph that compares the results with different training data sets

```{r}
  # Draw graph for 1 training sample

  MSE <- rep(0,10) # initialising the MSE vector (we will need it below when we add )
  
  set.seed(1) # initialize the randomizer with a different number
  train <- sample(nrow(Auto),nrow(Auto)/2) # generate indices of a training data set by random sampling half of the observation indices of Auto 
  AutoTrain <- Auto[train,] # training data 
  AutoTest <- Auto[-train,] # test data
  for (i in 1:10){
    lm.fit=lm(mpg~poly(horsepower,i),data=AutoTrain)
    MSE[i] <- mean((mpg-predict(lm.fit,Auto))[-train]^2)
  }
  
  x <- seq(1,10,1)
  plot(MSE~x, col=rainbow(1)[1], ylim=c(-2,30))
  lines(MSE~x, col=rainbow(1)[1])

  # Add graphs for different training samples
  for (j in 2:10){
    set.seed(j) # initialize the randomizer with different numbers
    train <- sample(nrow(Auto),nrow(Auto)/2) # generate indices of a training data set by random sampling half of the observation indices of Auto 
    AutoTrain <- Auto[train,] # training data 
    AutoTest <- Auto[-train,] # test data
    for (i in 1:10){
      lm.fit=lm(mpg~poly(horsepower,i),data=AutoTrain)
      MSE[i] <- mean((mpg-predict(lm.fit,Auto))[-train]^2)
  }
  
  points(MSE~x, col=rainbow(j)[j]) # add to the previous plot
  lines(MSE~x, col=rainbow(j)[j])
  
}
```

# Validation Set Approach (VSA)

*Advantages*
1.  Simple
2.  Easy to implement

*Disadvantages*
*1.  High varaince (low stability)* on some data sets.
*2.  High bias* on some data sets: since the training data is only a part of our sample, the VSA tends to overestimate the test error. We would prefer to train our model on the whole sample because statistical methods tend to perorm better the more observations they can use to learn.

# Exercise 2 - Leave-One_out Cross-Validation (LOOCV)

The LOOCV estimate can be automatically computed for any generalized linear model using the glm() together with the cv.glm() function. We‘ve used the glm() before for logistic regression by passing in the family="binomial" argument. But if we use glm() to fit a model without passing in the family argument, then it performs linear regression, just like the lm() function. 

```{r}
glm.fit <- glm(mpg~horsepower, data=Auto)
coef(glm.fit)

lm.fit <- lm(mpg~horsepower, data = Auto)
coef(lm.fit)
```
Both give the same values

Without speciying "family" as a parameter, glm() falls back to linear regression.
We use it here because it comes with a function for cross-validation.
```{r}
glm.fit <- glm(mpg~horsepower, data = Auto)

cv.err <- cv.glm(Auto,glm.fit) # computing the LOOCV prediction error. If we don't set the parameter K, cv.glm defaults to LOOCV
cv.err$delta[1] # the estimate for the test error is stored in delta[1]
```

Now we calculate the LOOC test error for polynomial regression models of degree 1, ..., 10.
It takes a bit to evaluate.
```{r}
cv.error <- rep(0,10) # initialising the LOOCV error vector
for (i in 1:10){
  glm.fit <- glm(mpg~poly(horsepower,i), data = Auto)
  cv.error[i] <- cv.glm(Auto,glm.fit)$delta[1]
}
cv.error
```

Plot the results
```{r}
x<- seq(1,10,1)
plot(cv.error~x, col="blue3")
lines(cv.error~x, col="blue3")
```

# k-Fold Cross-Validation

```{r}
set.seed(17)
```

Now we calculate the k-fld CV test error for polynomial regression models of degree 1,...,10.
Notice that the compution time is much shorter than that of LOOCV.
```{r}
cv.error.10=rep(0,10)
for (i in 1:10){
  glm.fit = glm(mpg~poly(horsepower,i),data=Auto)
  cv.error.10[i] = cv.glm(Auto,glm.fit,K=10)$delta[1] # Setting K=10 means 10-fold CV
}
cv.error.10
```
Plot the results
```{r}
plot(cv.error.10~x, col="blue")
lines(cv.error.10~x, col="blue")
```
